Supplement A to 'Researchers understand and use the logic of significance testing in a novel statistical reasoning task

### 2020-10-21 inst/manuscript/manuscript.Rmd
first draft of revision

The interpretation of studies of researchers' understanding of SST are limited, however, by their methodology. A typical study presents a vignette describing research results. Statistical results are offered to the participants (e.g., a $t$ statistic and $p$ value), who are then asked to explicitly give or endorse various interpretations that are thought to represent participants' understanding of SST. However, there are reasons to be cautious of drawing strong conclusions from these studies, including the abstract nature of such vignettes, the lack of investment researchers have in the fictional research, and their disconnection from research activity (e.g., experimentation and replication). It is unclear how well vignette studies (including ones by the present authors: [@Hoekstra:etal:2014]) tap understanding of the core logic of SST or, say, familiarity with the technical terminology used to present statistical results. Conceptual understanding and fluency with common representations are both important, but are distinct.

@@

Instead of the usual human-factors approach, which adopts familiar statistical language and tests participants' fluency with existing procedures, we adopt a more cognitive perspective. We aim to test working scientists' understanding of the basic conceptual framework underlying SST using a simulated experimental task. To avoid conflating basic reasoning processes and fluency with common statistical terminology, we avoid using common statistics---or, indeed, any numbers---at all.

@@

Participants were scientists or trainees recruited via social media. Our statistical reasoning task required them to perform a series of experiments to judge which of two groups of "Christmas elves" --- "Jinglies" or "Sparklies" --- could make more of a particular toy[^demo]. Because the study was run around the Christmas holiday season, it was hoped that the theme would make the task more enjoyable. The numerical information for an experiment, including sample size and the test statistic, was translated into color and location and displayed as a point on a two-dimensional visual interface (Figure \ref{fig:interface}). Participants could change the sample size per group for each experiment (increasing the time required to return a result), but did not know its numerical value. Importantly, the meaning of the colors and locations was unknown to the participants, aside from the monotone relationship with the sample size and statistical evidence.

[^demo]: A demonstration version of the task that does not store data can be found at https://richarddmorey.github.io/Morey_Hoekstra_StatCognition/articles/task_demo.html.

@@

Participants were randomly assigned to one of 15 effect size conditions: either no difference ($\delta=0$), or $\delta=\pm0.1$, $\pm0.185$, $\pm0.296$, $\pm0.433$, $\pm0.596$,$\pm0.785$, or $\pm1$ standard deviation units. Each participant had a 25% probability of being assigned $\delta=0$, with the other 75% being randomly and uniformly distributed across the remaining 14 effect size conditions. This true effect size was not revealed to them. Their goal was to determine the sign of the effect (ie, which of the two teams is truly faster).

Consistent with the fictional two-sample design, statistical evidence for each "experiment" was sampled from a normal distribution with mean that depended on the (chosen, but unknown) sample size and their randomly assigned effect size:
\[
Z \sim \mbox{Normal}(\delta\sqrt{n/2}, 1)
\]
This was mapped $Z$ test statistic was mapped into a horizontal location on the interface through an arbitrary function unknown to the participant. Participants were randomly assigned to one of two mapping functions: a "wide" function, and a narrow function (see Supplement`~A`{=latex}, section`~1.2`{=latex} for full mathematical details). Figures`~`{=latex}\ref{fig:interface}A and B, show the visual effect of this manipulation. Statistically, these two conditions were identical; visually, they were not.

@@

Participants sought out information that would be necessary for significance tests. They made heavy use of shuffle reports (Figure~{=latex}\@ref(fig:sampling)A). Across all true effect sizes, participants sampled a median of 152 shuffle reports (range: 1-2034; in both panels A and B, lines show robust regression fits [@Venables:Ripley:2002]).

@@

```{r error-rates, fig.cap="Correct decision and error rates as a function of true effect size. Positive and negative effect sizes hav been collapsed. The red solid line represents the rate at which participants {\\em correctly} determined the direction of the effect, except for at an effect size of 0, where it represents an incorrect claim of an effect. The shaded region around the solid line is one standard error wide. The difference between the black and the gray dashed lines shows the proportion of participants who decided that the two groups were the same (as opposed to failing to detect a difference).", fig.showtext=FALSE, out.width=".9\\columnwidth"}
MoreyHoekstra2019::fig_error_rates()
```

Decision error rates as a function of true effect size are shown in Figure \ref{fig:error-rates}.

@@

Of the `r null_n` participants for whom the null hypothesis was true (i.e. $\delta=0$), `r t1_err_count` participants (`r round(t1_err_count/null_n * 100, 1)`%) incorrectly indicated an effect. This is larger than the typically-accepted 5% false positive rate in many sciences; however, participants were performing a novel task with no recourse to numbers or statistical software. Those who did not indicate an effect when $\delta=0$ tended to indicate that they *did not detect* an effect (`r null_nodet_count`; `r round(null_nodet_count/null_n * 100,1)`%), which is the correct conclusion from the SST perspective. The other `r null_same_count` (`r round(null_same_count/null_n * 100,1)`%) indicated that the groups were the same, which under SST is a fallacy.

@@

Another way one can evaluate the participants' responses is whether they reflect the information in the display at the time the decision is made, taking into account all points. To roughly quantify the evidence for a difference for each participant, we computed two $p$ values from Wilcoxon tests using the fictitious experimental results as they stood when the participant made their decision: a signed-rank test on the experimental samples alone, and a rank-sum test between the shuffle reports and the experimental samples. These two $p$ values indicate the information available to participants using sign-like significance tests and those using the null samples, respectively. The rank-sum $p$ value was typically lower, but to fairly account for the information available to the participant, we took the lower of the two. Small Wilcoxon $p$ values suggest a difference between the shuffle reports and the experiments.

@@

Although it has previously been suggested that scientists have dramatic misunderstandings of SST logic, scientists and trainees in our experiment demonstrate both understanding and the ability to use it to come to the correct conclusion in a simulated statistical task. Moreover, they report strategies consistent with SST, and the signature of SST reasoning can be seen in their responses. Because we removed numerical effect size and sample size information --- making strategies other than pure significance testing difficult or impossible to apply --- our results are strong evidence that scientists *can* successfully deploy SST logic. It is still an open question what causes SST to be misunderstood so often, but it seems comprehension of its underlying logic can be ruled out for many scientists.

There are some major differences between our methods and the methods of previous demonstrations that scientists fail when reasoning about SST. We have already mentioned the fact that previous demonstrations typically use survey vignettes, as opposed to our more engaged experimental method. A second difference is our sample which is a substantially larger and more diverse convenience sample than previous demonstrations. It is possible that the recent so-called "replication crisis" [@Baker:2016] has raised awareness among scientists of deficits of statistical reasoning. If this increase in awareness accounts for our more optimistic findings, this would be good news for statistical educators.

Our findings echo other work showing that human reasoning can, under some conditions, be better than previously understood. [@Cosmides:Tooby:1992;@Gigerenzer:Hoffrage:1995]. Suggestions that SST be discontinued due to scientists' apparent misunderstandings may have been hasty. Of course, there may be other reasons why some favor abandoning SST, but our work shows that given the opportunity, scientists successfully deploy basic SST logic. In spite of scientists' real-life statistical behaviour often resembling a "ritual" [@Gigerenzer:etal:2004], when we eliminate the ritual --- no $p$ value, or any other familiar number, was offered --- they think statistically, very often arriving at the correct conclusion about the sign of the effect. Effective methodological reform in the sciences may be easier than abandoning significance testing, if we can harness scientists' already-existing statistical competence.

>>>>

The interpretation of studies of researchers' understanding of SST are limited, however, by their methodology. A typical study presents a vignette describing research results. Statistical results are offered to the participants (e.g., a $t$ statistic and $p$ value), who are then asked to explicitly give or endorse various interpretations that are thought to represent participants' understanding of SST. However, there are reasons to be cautious of drawing strong conclusions from these studies, including the abstract nature of such vignettes, the lack of investment researchers have in the fictional research, and their disconnection from research activity (e.g., experimentation and replication). It is unclear how well vignette studies (including ones by the present authors: [@Hoekstra:etal:2014;@Hoekstra:etal:2012]) tap understanding of the core logic of SST or, say, familiarity with the technical terminology used to present statistical results. Conceptual understanding and fluency with common representations are both important, but are distinct.

@@

To avoid conflating basic reasoning processes and fluency with common statistical terminology, we avoid using common statistics---or, indeed, any numbers---at all. Instead of focusing on familiar statistical language and tests participants' fluency with existing procedures, we adopt a different approach: we test working scientists' understanding of the basic conceptual framework underlying SST using a simulated experimental task.

@@

Participants were scientists or trainees recruited via social media. Our statistical reasoning task required them to perform a series of experiments to judge which of two groups of "Christmas elves" --- "Jinglies" or "Sparklies" --- could make more of a particular toy. A demonstration version of the task that does not store data can be found at https://richarddmorey.github.io/Morey_Hoekstra_StatCognition/articles/task_demo.html. Because the study was run around the Christmas holiday season, it was hoped that the theme would make the task more enjoyable. The numerical information for an experiment, including sample size and the test statistic, was translated into color and location and displayed as a point on a two-dimensional visual interface (Figure \ref{fig:interface}). Participants could change the sample size per group for each experiment (increasing the time required to return a result), but did not know its numerical value. Importantly, the meaning of the colors and locations was unknown to the participants, aside from the monotone relationship with the sample size and statistical evidence.

@@

Participants were randomly assigned to one of 15 effect size conditions: either no difference ($\delta=0$), or $\delta=\pm0.1$, $\pm0.185$, $\pm0.296$, $\pm0.433$, $\pm0.596$,$\pm0.785$, or $\pm1$ standard deviation units. Each participant had a 25% probability of being assigned $\delta=0$, with the other 75% being randomly and uniformly distributed across the remaining 14 effect size conditions. These true effect sizes were not revealed to the participants. Their goal was to determine the sign of the effect (ie, which of the two teams is truly faster).

Consistent with the fictional two-sample design, statistical evidence for each "experiment" was sampled from a normal distribution with mean that depended on the (chosen, but unknown) sample size and their randomly assigned effect size:
\[
Z \sim \mbox{Normal}(\delta\sqrt{n/2}, 1)
\]
The $Z$ test statistic was mapped into a horizontal location on the interface through an arbitrary function unknown to the participant. Participants were randomly assigned to one of two mapping functions: a "wide" function, and a narrow function (see Supplement`~A`{=latex}, section`~1.2`{=latex} for full mathematical details). Figures`~`{=latex}\ref{fig:interface}A and B show the visual effect of this manipulation. Statistically, these two conditions were identical; visually, they were not.

@@

Participants sought out information that would be necessary for significance tests. They made heavy use of shuffle reports (Figure`~A`{=latex}\@ref(fig:sampling)). Across all true effect sizes, participants sampled a median of 152 shuffle reports (range: 1-2034; in both panels A and B, lines show robust regression fits [@Venables:Ripley:2002]).

@@

```{r error-rates, fig.cap="Observed and fitted probabilities for each effect size and response. For negative and positive effect sizes, the correct response \"Sparklies\" or \"Jinglies\" respectively. Fitted probabilities are from the signal detection model outlined in Supplement A. Lines show predicted probabilities; ribbons show where 68\\% of the observed probabilities should fall given the predicted probabilities. These limits are approximate due to the discreteness of the response.", fig.showtext=FALSE, out.width=".9\\columnwidth"}
MoreyHoekstra2019::fig_error_rates_dprime()
```

Decision rates as a function of true effect size are shown in Figure \ref{fig:error-rates}.

@@

Of the `r null_n` participants for whom the null hypothesis was true (i.e. $\delta=0$), `r t1_err_count` participants (`r round(t1_err_count/null_n * 100, 1)`%) incorrectly indicated an effect. This is larger than the typically-accepted 5% false positive rate in many sciences; however, participants were performing a novel task with no recourse to numbers or statistical software. Those who did not indicate an effect when $\delta=0$ tended to indicate that they *did not detect* an effect (`r null_nodet_count`; `r round(null_nodet_count/null_n * 100,1)`%), which is the correct conclusion from the SST perspective. The other `r null_same_count` (`r round(null_same_count/null_n * 100,1)`%) indicated that the groups were the same, which under SST is typically considered a fallacy.

@@

Another way one can evaluate the participants' responses is whether they reflect the information in the display at the time the decision is made, taking into account all points. To roughly quantify the evidence for a difference for each participant, we computed two $p$ values from Wilcoxon tests using the fictitious experimental results as they stood when the participant made their decision: a signed-rank test on the experimental samples alone, and a rank-sum test between the shuffle reports and the experimental samples. These two $p$ values indicate the information available to participants using sign-like significance tests and those using the null samples, respectively. The rank-sum $p$ value was typically lower; although it makes little difference to the qualitative results, to fairly account for the information available to the participant, we used the smaller of the two $p$ values. In general, smaller $p$ values suggest a larger observed between the shuffle reports and the experiments, allowing us to compare the stimulus the participants were given to their decisions.

@@

Although it has previously been suggested that scientists have dramatic misunderstandings of SST logic, scientists and trainees in our experiment demonstrate both understanding and the ability to use it to come to the correct conclusion in a simulated statistical task. Moreover, they report strategies consistent with SST, and the signature of SST reasoning can be seen in their responses. Because we removed numerical effect size and sample size information --- making strategies other than pure significance testing difficult or impossible to apply --- our results are evidence that scientists *can* successfully deploy SST logic. It is still an open question what causes SST to be misunderstood so often, but it seems comprehension of its underlying logic can be ruled out for many scientists.

Our findings echo other work showing that human reasoning can, under some conditions, be better than previously understood. [@Cosmides:Tooby:1992;@Gigerenzer:Hoffrage:1995]. Suggestions that SST be discontinued due to scientists' apparent misunderstandings may have been hasty. Of course, there may be other reasons to abandon SST, but our work shows that given the opportunity, scientists successfully deploy basic SST logic. In spite of scientists' real-life statistical behaviour often resembling a "ritual" [@Gigerenzer:etal:2004], when we eliminate the ritual --- no $p$ value, or any other familiar number, was offered --- they think statistically, very often arriving at the correct conclusion about the sign of the effect.

We wish to emphasize what we cannot, and do not, argue. First, we cannot argue that simply because scientists can successfully use SST logic, that they do in real situations, or that specific instantiations of SST, such as $p$ values, are used well. We specifically set out to abstract the logic away from the typical ways and situations in which scientists use the logic. This has the benefit of being helping to identify where problems might be, but the downside that generalizing the results will require further work. We are also not addressing other potential arguments against SST, such as philosophical ones.

Although our conclusions are different those of other studies, there are some major differences between our methods and the methods of previous demonstrations that scientists fail when reasoning about SST. We have already mentioned the fact that previous demonstrations typically use survey vignettes, as opposed to our more engaged experimental method. A second difference is our sample which is a substantially larger and more diverse convenience sample than previous demonstrations. It is possible that the recent so-called "replication crisis" [@Baker:2016] has raised awareness among scientists of deficits of statistical reasoning. If this increase in awareness accounts for our more optimistic findings, this would be good news for statistical educators.

Finally, we hope to provide a fresh method and perspective on a long-standing debate in statistical cognition. Simulation-based approaches to teaching statistics have long been touted [@Cumming:etal:1995;@Rossman:Chance:2014]. Simulation-based approaches to *studying* scientists' statistical reasoning may also profitable, particularly in studying reasoning that is difficult for participants to articulate formally. If we are to reform statistical education and practice in the sciences, we should base that reform on diverse lines of evidence about scientists' reasoning. Harnessing scientists' already-existing statistical competence may make effective methodological reform in the sciences easier than wholesale abandoning of SST.

### 2020-10-21 inst/manuscript/supplementA.Rmd
first draft of revision

# Participant sampling behavior

```{r sampling, fig.cap="Sampling behavior by effect size. Each point represents a participant. A: Number of samples from the null distribution as a function of true effect size. B: Number of samples of fictional 'experiments' as a function of true effect size. Note that the $y$ axis is logarithmically scaled. Lines are robust regression fits (Venables \\& Ripley, 2002).", fig.showtext=FALSE}
MoreyHoekstra2019::fig_sampling_behavior()
```

```{r}
kt_null <- kruskal.test(n_null ~ factor(abs(effect_size)), data = MoreyHoekstra2019::christmas_stats_participants)
kt_expt <- kruskal.test(n_expt ~ factor(abs(effect_size)), data = MoreyHoekstra2019::christmas_stats_participants)
```

Participants sought out information that would be necessary for significance tests. They made heavy use of shuffle reports (Figure A above). Across all true effect sizes, participants sampled a median of 152 shuffle reports (range: 1-2034).

Participants also made use of "replications" of the fictional experiments. (Figure B above) shows the distribution of the number of experiments sampled as a function of the true effect size. Median numbers of experiments range from 20 when Jinglies and Sparklies were equally fast, down to 9 when the true effect size was $\delta=1$ and thus the effect was relatively easy to detect (`r papaja::apa_print.htest(kt_expt)$statistic`). When the effect size is small and difficult to detect, participants do more experimentation before deciding.

@@

>>>>

@@

# Signal detection analysis

We can treat the participants as "detectors" of the true effect, and ask how well the scientists were able to distinguish signal from noise, in the aggregate. To do this, we fit a signal detection model that allowed $d'$ to vary as a function of the true effect size.

Let $d'_i$ ($i=-7,\ldots,7$) be the means for the 15 effect size conditions from $\delta=-1$ to $\delta=1$. We constrain $d'_0=0$ , $d'_i\leq d'_j$ when $i<j$ (monotonicity), and $d_{i}=-d_{-i}$ (symmetry). The probabilities of responding "Jinglies" or "Sparklies" in the $i$th effect size condition are
\[
\begin{eqnarray}
Pr_i(\mbox{Sparklies}) &=& \Phi\left(c_1 - d'_i\right)\\
Pr_i(\mbox{Jinglies}) &=& 1 - \Phi\left(c_2 - d'_i\right)
\end{eqnarray}
\]
where $\Phi$ is the cumulative distribution function of the standard normal distribution and $c_1,c_2$ are criteria (not necessarily symmetric). The probability of responding "same" or "cannot detect" is the remaining probability.

The model has 9 parameters (two criteria and seven means, though the monotonicity assumption adds more constraint) for 30 non-redundant data points. The model was fit using maximum likelihood (R code is available in the source of this document).

```{r sigdet1, fig.cap="Aggregate signal detection analysis. Fitted d' for each effect size. See text for a description of the fitted model.",fig.showtext=FALSE}

MoreyHoekstra2019::christmas_stats_participants %>%
  mutate(
    response = as.character(response),
    response = case_when(
      response %in% c("same","no_detect") ~ "no detect/same",
      TRUE ~ response
    )
  ) %>%
  group_by(effect_size, response) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = "response", values_from = "n") %>%
  replace_na(list("no detect/same"=0, "sparklies"=0, "jinglies"=0)) %>%
  MoreyHoekstra2019:::sigdet_model_fit() -> sd_results

dp = sd_results$dp

MoreyHoekstra2019::fig_dprime_effect_size()

Figure \@ref(fig:sigdet1) shows the estimated $d'$ parameters as a function of effect size. They range from `r round(dp[9],3)` when $\delta=.1$ to `r round(dp[15],3)` when $\delta=1$.

```{r sigdet2, fig.cap = "Observed and fitted probabilities for each effect size and response. Lines show predicted probabilities; ribbons show where 68% of the observed probabilities should fall given the predicted probabilities. These limits are approximate due to the discreteness of the response.",fig.showtext=FALSE}

MoreyHoekstra2019::fig_error_rates_dprime()
```

 Figure \@ref(fig:sigdet2) shows the predicted and observed probabilities of all responses. The good fit of this model licenses the interpretation of the $d'$ parameters and the collapsing across the sign of the effect in Figure 3 of the main document.
